Formalizing the Kolmogorov extension Theorem in Lean

### 2023-09-14 Manuscript/proof_outline.tex
rename todo in lemma name

%\todo[inline]{Clearly differentiate refs to mathlib vs refs to new lemmas?}

@@

%\todo[inline]{explain leanlinem}

@@

At first sight, it might be surprising that martingales, a certain class of stochastic process, have already been formalized in Lean \cite{ying2023formalization}, although the Kolmogorov extension theorem (on the existence of stochastic processes) is only available by our contribution. Note, however, that a martingale $(X_t)_{t\in \iota}$ is defined as a family of random variables (satisfying some properties) on a fixed probability space $(\Omega, \mathcal F, \mathbf P)$, while the Kolmogorov extension theorem is on the construction of a probability (or finite) measure $\mathbf P$, on which we can define random variables (with certain properties). So, our work complements \cite{ying2023formalization} in the sense that probability spaces $(\Omega, \mathcal F, \mathbf P)$ on which you can define a martingale exist. {\color{blue}As an application of our implementation, we construct a probability measure on an infinite product space where all coordinates are independent; see Section~\ref{ss:concl}.} % However, we do not give any specific example for the construction of a specific instance of such a space\todo{could we give one?}.

The main steps in our construction, which were previously missing in {\tt mathlib}, are (more mathematical details are below, and in standard textbooks on Probability Theory, e.g.\ \cite{Klenke2013, Kallenberg2020}):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see
  Definition~\ref{def:semi});
\item {\color{blue}a definition of additive contents over semi-rings (see Definition~\ref{def:content});}
%\item adding some main lemmas to measure theory, e.g.\ continuity of a measure at the emptyset\todo{for a measure, this is already in mathlib?} (see e.g.\ Lemma~\ref{l:halbextRing}, Proposition~\ref{P:stetigmass}, Lemma~\ref{l:stetigcompact} below);
\item a proof that a single probability measure on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg}; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us with a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps (See Theorem~\ref{T1}).
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. This formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical. In addition, {\color{blue}Isabelle works with simple type theory, while Lean is an implementation of dependent theory.} %However, type theory as the basis of the formalization, as well as several mathematical concepts (e.g.\ projective families of measures or the continuity of a measure in the emptyset) are the same in both formalisations\todo{no, the type theory is not the same} in these two interactive theorem provers.

@@

% \todo{it would be nice to explain how the formalization led us to
% remove that hypothesis}.

@@

From the field/ring/semi-ring of cylinders, we have to define the
generated $\sigma$-algebra. This uses the following:
%\todo{one of those 2 is ours, the other is from mathlib. How do we show the  difference?}:
\begin{minted}[highlightlines={1}, framesep = 0pt, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  theorem generateFrom_cylinders :
    MeasurableSpace.generateFrom (cylinders α)
    = MeasurableSpace.pi

@@

\noindent
The existence-statement of the above lemma actually gives rise to a
definition. Here, we use \leanline{Exists.choose} in order to extract
an element from an $\exists$-statement. In addition, we do not allow
$\emptyset \in \mathcal K$ without loss of generality\todo{that
  exclusion was useful for something. Find out what that is}:

@@

\todo[inline]{Rename the todo! \xout{And replace by the version about contents?}}

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  theorem countably_additive_addContent_of_todo
  (hC : SetRing C)
  (m : AddContent C)
  (hm_ne_top : ∀ {s} (_ : s ∈ C), m s ≠ ∞)
  (hm : ∀ ⦃s : ℕ → Set α⦄ (_ : ∀ n, s n ∈ C),
    Antitone s → (⋂ n, s n) = ∅ →
    Tendsto (fun n => m (s n)) atTop (nhds 0))
    ⦃f : ℕ → Set α⦄ (hf : ∀ i, f i ∈ C) (hUf : (⋃ i, f i) ∈ C)
    (h_disj : Pairwise (Disjoint on f)) :
  m (⋃ i, f i) = ∑' i, m (f i) 
\end{minted}

>>>>

@@

@@

At first sight, it might be surprising that martingales, a certain class of stochastic process, have already been formalized in Lean \cite{ying2023formalization}, although the Kolmogorov extension theorem (on the existence of stochastic processes) is only available by our contribution. Note, however, that a martingale $(X_t)_{t\in \iota}$ is defined as a family of random variables (satisfying some properties) on a fixed probability space $(\Omega, \mathcal F, \mathbf P)$, while the Kolmogorov extension theorem is on the construction of a probability (or finite) measure $\mathbf P$, on which we can define random variables (with certain properties). So, our work complements \cite{ying2023formalization} in the sense that probability spaces $(\Omega, \mathcal F, \mathbf P)$ on which you can define a martingale exist. {\color{blue}As an application of our implementation, we construct a probability measure on an infinite product space where all coordinates are independent; see Section~\ref{ss:concl}.}

The main steps in our construction, which were previously missing in {\tt mathlib}, are (more mathematical details are below, and in standard textbooks on Probability Theory, e.g.\ \cite{Klenke2013, Kallenberg2020}):
\begin{enumerate}
\item a formalization of (the set-system of) semi-rings (see
  Definition~\ref{def:semi});
\item {\color{blue}a definition of additive contents over semi-rings (see Definition~\ref{def:content});}
\item a proof that a single probability measure on a Polish space is inner regular with respect to compact sets (see Definition~\ref{def:innerreg}; in fact, we proved a slight generalization using pseudo-metric spaces, found in Lemma~\ref{L:relcoPol} below); 
\item the classical Carathéodory extension theorem, providing us with a candidate for the measure which we want to construct (see Theorem~\ref{T:masseind}); 
\item the Kolmogorov extension theorem, as based on the previous steps (See Theorem~\ref{T1}).
\end{enumerate}
The Kolmogorov extension theorem has been previously formalized in Isabelle/HOL \cite{Immler2012}. This formalization only works on Polish spaces (rather than on spaces where every finite measure is inner regular with respect to compact sets, see below), and only in the case where all $\alpha_i$'s are identical. In addition, {\color{blue}Isabelle works with simple type theory, while Lean is an implementation of dependent theory.}

@@

@@

From the field/ring/semi-ring of cylinders, we have to define the
generated $\sigma$-algebra. This uses the following:
\begin{minted}[highlightlines={1}, framesep = 0pt, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  theorem generateFrom_cylinders :
    MeasurableSpace.generateFrom (cylinders α)
    = MeasurableSpace.pi

@@

\noindent
The existence-statement of the above lemma actually gives rise to a
definition. Here, we use \leanline{Exists.choose} in order to extract
an element from an $\exists$-statement. In addition, we do not allow
$\emptyset \in \mathcal K$ without loss of generality:

@@

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
  theorem sigma_additive_addContent_of_tendsto_zero
  (hC : SetRing C)
  (m : AddContent C)
  (hm_ne_top : ∀ {s} (_ : s ∈ C), m s ≠ ∞)
  (hm : ∀ ⦃s : ℕ → Set α⦄ (_ : ∀ n, s n ∈ C),
    Antitone s → (⋂ n, s n) = ∅ →
    Tendsto (fun n => m (s n)) atTop (nhds 0))
    ⦃f : ℕ → Set α⦄ (hf : ∀ i, f i ∈ C) (hUf : (⋃ i, f i) ∈ C)
    (h_disj : Pairwise (Disjoint on f)) :
  m (⋃ i, f i) = ∑' i, m (f i) 
\end{minted}
